\begin{abstract}
While the National Institute of Standards and Technology (NIST) has now finalized the first set of PQC standards in August 2024 \cite{fips203, fips204, fips205}, a glaring omission has been the lack of cross-algorithm benchmarking for all three classes of constrained devices defined by RFC 7228 for IoT devices. This paper provides a comprehensive empirical analysis of the five key PQC algorithms: ML-KEM (KYBER), ML-DSA (DILITHIUM), FN-DSA (FALCON), SLH-DSA (SPHINCS+), and NTRU, over fourteen distinct parameter sets covering all three classes of constrained devices: Class 0, Class 1, and Class 2. Through over 45,000 rigorous empirical approximations based on bare-metal instruction limits, this work examines execution time, memory size, energy consumption, and communication overheads. An enhanced energy estimation framework is proposed, accounting for both processor cycles for crypto functions and communication latency for transmission functions. In addition, a TOPSIS-based decision framework is proposed to guide algorithm selection based on multiple criteria. Statistical modeling, validated through ANOVA and Tukey HSD significance testing (all results significant at $p < 0.001$), has identified key performance bottlenecks for two key PQC algorithms: ML-DSA (DILITHIUM)-2, which requires $>50$ KB peak memory, thereby leading to severe Out-of-Memory (OOM) failures on Class 1 devices, and SLH-DSA (SPHINCS+)-128f, where execution latencies $>540$ ms impose key limitations on real-time capabilities. We formally demonstrate that Class 0 devices ($< 10$ KB RAM) are entirely infeasible for direct PQC integrations without symmetric AES-GCM gateway delegation architectures. Conversely, ML-KEM (KYBER)-512 operating alongside FN-DSA (FALCON)-512 (requiring hardware FPU) provides the optimal Pareto-efficient trade-off for Class 1 networks. We conclude by projecting simulated TLS 1.3 handshake latencies (averaging 19.3 ms for Class 2 hybrids), assessing hybrid classical-quantum transition overheads (triggering 200\% computational loads), and outlining a definitive enterprise migration roadmap.
\end{abstract}
