%-----------------------------------------------------------------------
\section{Threat Model and Security Assumptions}
\label{sec:threat_model}
%-----------------------------------------------------------------------

To contextualize the urgency and scope of IoT migration towards post-quantum standards, we formally define our assumed threat geometry governing resource-constrained interactions.

\subsection{Quantum Adversarial Model}

Our adversary, $\mathcal{A}_{\text{CRQC}}$, is defined as a non-local, passive and active networked entity possessing full intercept, alter, and rerouting capabilities corresponding strictly to the standard Dolev-Yao threat model applied over long-range IoT transmission spectrums (e.g., LoRaWAN, NB-IoT, or wide-area Wi-Fi).

Crucially, $\mathcal{A}_{\text{CRQC}}$ possesses computational access to a cryptographically relevant quantum computer executing a hybridized combination of Grover's Algorithm \cite{grover1996fast} targeting symmetric collisions and Shor's Algorithm \cite{shor1994algorithms} targeting Abelian hidden subgroup problems spanning integer factorization and elliptic curve discrete logarithms. We assume $\mathcal{A}_{\text{CRQC}}$ possesses sufficient stable logical qubits required to execute arbitrary-depth quantum circuitry to breach NIST Security Category 1 ($\geq$ 128-bit classical equivalent) systems instantly.

\subsection{The "Harvest Now, Decrypt Later" Immediacy}

Unlike physical infrastructure attacks, cryptographic breaches inherently traverse time. IoT traffic streams---such as digitized medical telemetry logs or encrypted municipal grid commands---are captured indiscriminately as bulk ciphertexts today. Because the long-term confidentiality thresholds of these records frequently exceed 25 years, the retroactive exposure hazard effectively demands present-day PQC adoption regardless of immediate commercial quantum availability metrics.

\subsection{Side-Channel Vulnerabilities}

Beyond purely algebraic quantum breaches, $\mathcal{A}_{\text{CRQC}}$ is assumed to possess extensive localized hardware penetration metrics. Specifically:
\begin{itemize}
    \item \textbf{Timing Attacks:} Adversaries monitor minor CPU execution variations occurring during non-constant-time polynomial branch logic (e.g., Gaussian rejection sampling bounds typical in \falcon{}).
    \item \textbf{Differential Power Analysis (DPA):} Adversaries infer the secret lattice error term polynomial gradients strictly by correlating power trace consumption spikes executed by the IoT microcontroller mapping instructions to memory \cite{kocher1999differential, hutter2011side}.
\end{itemize}
These models heavily mandate that deployed IoT solutions execute strictly constant-time functions. Constant-time algorithms unconditionally execute at identical clock-cycle bounds regardless of input payloads, deliberately neutralizing external timing leakages at the explicit cost of massive performance saturation.
