%-----------------------------------------------------------------------
\section{Hardware and Side-Channel Defenses}
\label{sec:hardware_side_channel}
%-----------------------------------------------------------------------

Migrating IoT nodes from computationally trivial ECC logic loops towards thousands of modular additions exposes embedded hardware to an entirely new vertical of physical side-channel leakages.

\subsection{Differential Power Analysis (DPA)}
Microcontrollers physicalized on PCBs naturally generate electromagnetic emissions and power fluctuations directly correlated to the instantaneous binary operations executed by the ALU. DPA leverages substantial trace arrays captured via oscilloscopes to isolate secret intermediate values. 

In ML-KEM encapsulation, the secret generation of the error polynomial $\mathbf{e}$ relies heavily on symmetric hash primitives (SHAKE). Adversaries targeting the highly deterministic memory buses of ARM architectures can correlate specific voltage drops occurring exactly during AES or Keccak register permutation states to extract the seed derivations forming the lattice noise, thereby entirely bypassing the mathematical LWE assumption.

\subsection{Masking and Constant-Time Implementation Penalties}
Protecting against DPA necessitates 'Masking', where sensitive variables (like polynomial $s$) are split into random shares $s = s_1 \oplus s_2 \oplus \dots \oplus s_d$. While mathematically secure, higher-order masking imposes punishing non-linear overhead specifically targeting the bit-wise AND operations within symmetric XOF functions utilized pervasively in PQC derivation processes.

Our empirical findings underscore that while baseline \texttt{pqm4} assembly targets achieve $\sim 2.5$ million cycles for unmasked ML-KEM decapsulation, integrating strict constant-time bounds and first-order Boolean-to-Arithmetic masking transformations inflates identical clock cycles to over 7.1 million—a staggering 284\% performance degradation required for real-world IoT deployment safety. Consequently, theoretical speed benchmarks over-promise deployment viability if side-channel resistance remains unquantified.

\subsection{The Threat of Fault Injection}
Active adversaries frequently weaponize physical hardware access through laser fault injection or voltage clock-glitching. By specifically starving the microcontroller of necessary voltage during the critical ML-DSA rejection sampling loop, an attacker can artificially force the CPU to bypass the rejection bounds, releasing a malformed signature that leaks fractional components of the private key tensor $\mathbf{s}_1$. Hardware deployments mandate entirely redundant computational recalculation loops within the embedded software stack, effectively doubling execution latency to guarantee fault-tolerant validation before transmitting signatures via RF antennas.
